SCPCL000824 - Validation of tumor cells

### 2024-06-17 analyses/cell-type-ewings/cell_type_annotations/SCPCS000492/SCPCL000824_tumor-cell-validation.Rmd
use saved functions

```{r}
# Input files
sce_file <- file.path(sample_dir, "SCPCL000824_processed.rds")
marker_genes <- file.path(module_base, "references", "tumor-marker-genes.tsv")

@@

```{r}
# read in all files 
sce <- readr::read_rds(sce_file)

# read in marker genes table 
marker_genes_df <- readr::read_tsv(marker_genes) |> 
  # account for genes being from multiple sources
  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |> 
  dplyr::distinct()

marker_gene_results_df <- readr::read_tsv(marker_gene_results_file)

# copykat output
copykat_results_df <- readr::read_tsv(copykat_predictions_file) |> 
  # only select no_ref classification and rename to be consistent with naming of other classifications
  dplyr::select(barcodes, copykat_classification = no_ref) |> 
  dplyr::mutate(
    copykat_classification = dplyr::case_when(
      copykat_classification == "aneuploid" ~ "Tumor",
      copykat_classification == "diploid" ~ "Normal",
      TRUE ~ copykat_classification
    )
  )

# infercnv output 
infercnv_results_df <- readr::read_tsv(infercnv_predictions_file)
infercnv_metadata_df <- read.delim(infercnv_metadata_file) |> 
  tibble::rownames_to_column("barcodes")

# remove MT and GL chr, comments in docs recommend that these are not useful and can be inaccurate 
chr_columns_to_keep <- which(!stringr::str_detect(names(infercnv_metadata_df), "chrMT|chrGL"))
infercnv_metadata_df <- infercnv_metadata_df[, chr_columns_to_keep]

# gene set scores 
geneset_scores_df <- readr::read_tsv(geneset_scores_file)
```

```{r}
# grab all coldata and add in all classifications and results to use for plotting throughout notebook 
# get list of marker genes to use 
marker_genes <- marker_genes_df |> 
  dplyr::filter(cell_type == "tumor") |> 
  dplyr::pull(ensembl_gene_id)

# get the gene expression counts for all marker genes
marker_gene_exp <- logcounts(sce[marker_genes, ]) |>
  as.matrix() |>
  t() |>
  as.data.frame() |>
  tibble::rownames_to_column("barcodes")

# pull out the UMAP coordinates and make a data frame to use for plotting
classification_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
      ) |> 
  # add in marker gene expression to dataframe 
  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
  # combine all genes into a single column for easy faceting
  tidyr::pivot_longer(
    cols = starts_with("ENSG"),
    names_to = "ensembl_gene_id",
    values_to = "gene_expression"
  ) |> 
  # join with marker gene df to get gene symbols for plotting 
  dplyr::left_join(marker_genes_df, by = c("ensembl_gene_id")) |>
  # get rid of excess columns
  dplyr::select(barcodes, UMAP1, UMAP2, gene_symbol, ensembl_gene_id, gene_expression) |> 
  # add in marker gene classifications
  dplyr::left_join(marker_gene_results_df) |> 
  # add in copykat results 
  dplyr::left_join(copykat_results_df) |> 
  # add in infercnv results and metadata
  dplyr::left_join(infercnv_results_df) |> 
  dplyr::left_join(infercnv_metadata_df) |> 
  # gene set scores 
  dplyr::left_join(geneset_scores_df)
```

@@

>>>>

```{r}
# source in helper functions: plot_gene_heatmap() and plot_cnv_heatmap() 
# create_classification_df() and create_marker_gene_df()
validation_functions <- file.path(module_base, "scripts", "utils", "tumor-validation-helpers.R")
source(validation_functions)

# source in helper functions: make_jaccard_matrix() and jaccard() 
jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
source(jaccard_functions)
```

```{r}
# Input files
sce_file <- file.path(sample_dir, "SCPCL000824_processed.rds")
marker_genes_file <- file.path(module_base, "references", "tumor-marker-genes.tsv")

@@

```{r}
# read in sce file 
sce <- readr::read_rds(sce_file)
```

```{r}
# generate classification df 
classification_df <- create_classification_df(
  sce,
  marker_gene_results_file,
  copykat_predictions_file, 
  infercnv_predictions_file,
  infercnv_metadata_file,
  geneset_scores_file
)
```

@@

```{r}
# plot tumor cells for each classification method 
geneset_plot_df <- classification_df |> 
  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification"), starts_with("scaled_mean")) |> 
  tidyr::pivot_longer(
    cols = starts_with("scaled_mean"),
    names_to = "geneset",
    values_to = "mean_score"
  ) |> 
  dplyr::mutate(
    geneset = stringr::word(geneset, -1, sep = "-")
  )

ggplot(geneset_plot_df, aes(x = UMAP1, UMAP2, color = mean_score)) +
  geom_point(size = 0.5, alpha = 0.5) +
  facet_wrap(vars(geneset)) +
  scale_color_viridis_c()
```

Add geneset classification

```{r}
# any cell that has > 0 expression of at least one geneset 
classification_df <- classification_df |>
  dplyr::mutate(
    geneset_classification = dplyr::if_else(
      `scaled_mean-RIGGI` > 0 & `scaled_mean-ZHANG` > 0 & `scaled_mean-SILIGAN` > 0,
      "Tumor",
      "Normal"
    ))
```

Compare gene set classification to marker gene classification 
```{r}
classification_columns <- c(
  "copykat_classification",
  "cnv_proportion_classification",
  "geneset_classification"
)

# create jaccard matrices for SingleR and CellAssign compared to tumor/normal 
jaccard_matrices <- classification_columns |>
  purrr::map(\(name) {
    make_jaccard_matrix(
      classification_df,
      "marker_gene_classification",
      name
    )
  }) |> 
  purrr::set_names("CopyKAT", "InferCNV", "Gene set")
```

```{r}
# Set heatmap padding option
heatmap_padding <- 0.2
ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(heatmap_padding, "in"))

# list of heatmaps looking at SingleR/ CellAssign vs tumor/normal 
heatmap <- jaccard_matrices |>
  purrr::imap(
    \(classification_mat, classification_method) {
      ComplexHeatmap::Heatmap(
        t(classification_mat), # transpose because matrix rows are in common & we want a vertical arrangement
        col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
        border = TRUE,
        ## Row parameters
        cluster_rows = TRUE,
        row_title = classification_method,
        row_title_gp = grid::gpar(fontsize = 12),
        row_title_side = "left",
        row_names_side = "left",
        row_dend_side = "right",
        row_names_gp = grid::gpar(fontsize = 10),
        ## Column parameters
        cluster_columns = FALSE,
        column_title = "Marker genes",
        column_title_gp = grid::gpar(fontsize = 12),
        column_names_side = "bottom",
        column_names_gp = grid::gpar(fontsize = 10),
        column_names_rot = 90,
        ## Legend parameters
        heatmap_legend_param = list(
          title = "Jaccard index",
          direction = "vertical",
          legend_width = unit(1.5, "in")
        ),
        show_heatmap_legend = classification_method == "CopyKAT",
      )
    }) |>
  # concatenate vertically into HeatmapList object
  purrr::reduce(ComplexHeatmap::`%v%`) |>
  ComplexHeatmap::draw(
    heatmap_legend_side = "right",
    # add a margin to the heatmap so labels don't get cut off
    padding = unit(c(2, 20, 2, 2), "mm")
  )
```

Add in classifications for geneset vs. CNV

```{r}
classification_df <- classification_df |>
  dplyr::mutate(
    exp_classification = dplyr::case_when(
      marker_gene_classification == "Normal" & geneset_classification == "Normal" ~ "Normal",
      marker_gene_classification == "Tumor" & geneset_classification == "Tumor" ~ "Tumor",
      .default = "Ambiguous"
    ), 
    all_cnv_classification = dplyr::case_when(
      copykat_classification == "Normal" & cnv_proportion_classification == "Normal" ~ "Normal",
      copykat_classification == "Normal" & cnv_proportion_classification == "Reference"~ "Normal",
      copykat_classification == "Tumor" & cnv_proportion_classification == "Tumor" ~ "Tumor",
      .default = "Ambiguous"
    )
  )
```

```{r}
geneset_plot_df |> 
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  ggplot(aes(x = mean_score, color = classification)) +
  geom_density(bw = 0.05) +
  facet_grid(rows = vars(method), 
             cols = vars(geneset))
```

```{r}
# generate marker genes df
plot_markers_df <- create_marker_gene_df(
  classification_df, 
  marker_genes_file
)

# calculate sum of the scaled gene expression values for plotting 
marker_density_df <- plot_markers_df |>
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  dplyr::mutate(
    method = dplyr::case_when(
      method == "marker_gene_classification" ~ "Marker genes only",
      method == "copykat_classification" ~ "CopyKAT - no reference",
      method == "cnv_proportion_classification" ~ "InferCNV proportion",
      method == "exp_classification" ~ "Expression combined",
      method == "all_cnv_classification" ~ "CopyKAT and InferCNV",
      method == "geneset_classification" ~ "EWS-FLI1 Gene set"
    )
  )

ggplot(marker_density_df, aes(x = sum_exp, color = classification)) +
  geom_density() +
  facet_wrap(vars(method))
```

```{r}
ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = geneset_classification)) +
  geom_point(size = 0.5, alpha = 0.5)
```

CNV heatmap annotated by copykat and marker gene prediction, two annotation columns

```{r}
annotation_df <- classification_df |> 
  dplyr::select(barcodes, ends_with("classification")) |> 
  unique()

# need to make a new annotation, since we want to annotate the rows here 
annotation <- ComplexHeatmap::rowAnnotation(exp_classification = annotation_df$exp_classification,
                                            all_cnv_classification = annotation_df$all_cnv_classification)

# plot gains in cnv
plot_cnv_heatmap(classification_df,
                 cnv_col = "proportion_scaled_dupli",
                 annotation = annotation,
                 legend_title = "Copy number \ngain")
```

```{r}
# plot gains in cnv
plot_cnv_heatmap(classification_df,
                 cnv_col = "proportion_scaled_loss",
                 annotation = annotation,
                 legend_title = "Copy number \nloss")
```

Density plot with mean proportion CNV from InferCNV on x-axis, colored by prediction, faceted by classification method

```{r}
classification_df |> 
  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification"), mean_cnv_proportion) |> 
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  ggplot(aes(x = mean_cnv_proportion, color = classification)) +
  geom_density(bw = 0.05) +
  facet_grid(rows = vars(method))
```

Mean InferCNV proportion on x-axis, mean copyKAT proportion on y-axis, color by scaled sum of marker gene expression

```{r}
cnv_prop_df <- classification_df |> 
  dplyr::select(barcodes, ends_with("classification"), starts_with("scaled_mean"), mean_cnv_proportion, mean_cnv_detection) |> 
  tidyr::pivot_longer(
    cols = starts_with("scaled_mean"),
    names_to = "geneset",
    values_to = "mean_score"
  ) |> 
  dplyr::mutate(
    geneset = stringr::word(geneset, -1, sep = "-")
  )

ggplot(cnv_prop_df, aes(x = mean_cnv_proportion, y = abs(mean_cnv_detection), color = mean_score)) +
  geom_point(size = 0.5, alpha = 0.5) +
  facet_wrap(vars(geneset)) +
  scale_color_viridis_c()
```

```{r}
ggplot(cnv_prop_df, aes(x = mean_score, y = abs(mean_cnv_detection))) +
  geom_point(size = 0.5, alpha = 0.5) +
  facet_wrap(vars(geneset)) +
  scale_color_viridis_c()
```

faceted density plot with marker gene expression on x axis, colored by copykat and infercnv prediction and faceted by marker gene

```{r}
geneset_scores_df |> 
  dplyr::pivot

ggplot(geneset_scores_df, aes(x = mean_score, color = ))
```

use SingleR to classify as tumor/normal using SCPCS000490

correlation of geneset score and marker gene score 
```{r}
geneset_plot_df |>

  dplyr::left_join(marker_density_df) |> 
  ggplot(aes(x = sum_exp, y = mean_score)) +
    facet_wrap()
```
